---
title: "Finding_Sig_Genes_human"
author: "Connor Gagen, Julia Keefe, Paloma Peters, Anna Mellizo Kroll, Anna Broerman"
date: "2025-04-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
library(knitr)
```

## Importing counts values
```{r import counts from Salmon}
# This loads in the counts from the NF core run
counts_matrix_human <- read.table("//scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/Project2_files/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

# This will prepare the counts for DESEQ by creating g2s
g2s_human <- data.frame(
  gene_id_human = rownames(counts_matrix_human),
  gene_name_human = counts_matrix_human[, 1]
)
# This removes the gene name from the counts matrix such that it is only numeric values
counts_matrix_human <- counts_matrix_human[ , -1]
# Round the values within the counts matrix. 
counts_integer_human <- round(counts_matrix_human)

```

## Creating sample sheet for DESeq from counts file
```{r creating DESeq sample sheet}
# This creates the data frame with the same columns as the counts_matrix
deseq_samples_human <- data.frame(
  sample_id_human = colnames(counts_matrix_human))
# Splits the sample_id to create a columns for time point and replicate columns
split_values_human <- strsplit(deseq_samples_human$sample_id_human, "_")
# This extracts the time point and the replicate info
time_values_human <- sapply(split_values_human, function(x) x[[2]])
replicate_values_human <- sapply(split_values_human, function(x) x[[3]])
# This will place the time point and replicate into the data frame.
deseq_samples_human$time_point_human <- time_values_human
deseq_samples_human$replicate_human <- replicate_values_human
# This converts the time point and replicate data into a factor.
deseq_samples_human$time_point_human <- factor(deseq_samples_human$time_point_human)
deseq_samples_human$replicate_human <- factor(deseq_samples_human$replicate_human)
```

## Running DESEQ2
```{r running deseq2}
# Test to see if the sample sheet matches the corresponding deseq_samples data frame
stopifnot(all(colnames(counts_integer_human) == deseq_samples_human$sample_id_human))
rownames(deseq_samples_human) <- deseq_samples_human$sample_id_human
# Sets up the dds
dds_time_point_human <- DESeqDataSetFromMatrix(countData = counts_integer_human,
                              colData = deseq_samples_human,
                              design = ~ time_point_human)
# Runs the DESeqq
dds_time_point_human <- DESeq(dds_time_point_human)
```

## Extracting results for each time point from DESEQ dds
```{r compile DESEQ2 results}

# retrieves the names corresponding to the different time points and Removes the first entry from deseq 
results_names_human <- resultsNames(dds_time_point_human)[-1]

# Setting up data frame 
res_df_human <- data.frame("gene_id_human" = character(), 
                     "baseMean" = numeric(), 
                     "log2FoldChange" = numeric(), 
                     "lfcSE" = numeric(),
                     "stat" = numeric(),
                     "pvalue" = numeric(),
                     "padj" = numeric(),
                     "gene_name_human" = character(),
                     "result_name_human" = character())

# run for loop to retreive results from all time comparisons in dds
for(i in 1:length(results_names_human)) {
  results_name_human <- results_names_human[i]
  res_human <- results(dds_time_point_human, name = results_name_human)
  tmp_res_df_human <- res_human %>% 
    as.data.frame() %>%
    rownames_to_column("gene_id_human") %>%
    merge(g2s_human) %>%
    mutate(result_name_human = results_name_human)
  res_df_human <- bind_rows(res_df_human, tmp_res_df_human)
  
}

```

##  Filtering to significant results P <0.05 and LFC abs(1)
```{r filtering DESEQ results to LFC > abs(1)& P <0.05}

filtered_res_df_human <- res_df_human %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

unique_sig_genes_human <- unique(filtered_res_df_human$gene_id_human)

output_dir_human <- "/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/Project2_files"

# Write the filtered DESeq2 results to CSV
write.csv(unique_sig_genes_human, file.path(output_dir_human, "unique_sig_genes_human.csv"), row.names = FALSE)

```
