```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
```

# Goal: script our RNAseq analysis pipeline to find genes changing significant across time from dox exposure. 

# Importing counts values
```{r import counts from Salmon}
# laoding Salmon counts from NF_CORE run 3.14
counts_matrix <- read.table("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/data/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

# preparing counts for DESEQ
# create g2s
g2s <- data.frame(
  gene_id = rownames(counts_matrix),
  gene_name = counts_matrix[, 1]
)
# removing gene_name column for analyses
counts_matrix <- counts_matrix[ , -1]
# Round counts to integer mode required for DESEQ2
counts_integer <- round(counts_matrix)

```

# Creating sample sheet for deseq from counts file
```{r creating deseq sample sheet}
# setting up data frame with cols from counts matrix
deseq_samples <- data.frame(
  sample_id = colnames(counts_matrix))
# creating time point and replicate cols from col names
# Split names
split_values <- strsplit(deseq_samples$sample_id, "_")
# grabbing time point and replicate info 
time_values <- sapply(split_values, function(x) x[[2]])
replicate_values <- sapply(split_values, function(x) x[[3]])
# placing time point and replicate info into data frame
deseq_samples$time_point <- time_values
deseq_samples$replicate <- replicate_values
# Factor time point and replicate
deseq_samples$time_point <- factor(deseq_samples$time_point)
deseq_samples$replicate <- factor(deseq_samples$replicate)

```

# Running DESEQ2
```{r running deseq2}
# testing sample sheet and counts are arranged properly 
stopifnot(all(colnames(counts_integer) == rownames(deseq_samples$sample_id)))

# setting up dds
dds_time_point <- DESeqDataSetFromMatrix(countData = counts_integer,
                              colData = deseq_samples,
                              design = ~ time_point)
# Run time point model
dds_time_point <- DESeq(dds_time_point)

```

# Extracting results for each time point from DESEQ dds
```{r compile DESEQ2 results}

# setting up result names object for input to forloop
result_names <- resultsNames(dds_time_point)
results_names <- result_names[-1]

# Setting up dataframe
res_df <- data.frame("gene_id" = character(), 
                     "baseMean" = numeric(), 
                     "log2FoldChange" = numeric(), 
                     "lfcSE" = numeric(),
                     "stat" = numeric(),
                     "pvalue" = numeric(),
                     "padj" = numeric(),
                     "gene_name" = character(),
                     "result_name" = character())

# run for loop to retreive results from all time comparisons in dds
for(i in 1:length(results_names)) {
  results_name <- results_names[i]
  res <- results(dds_time_point, name = results_name)
  tmp_res_df <- res %>% 
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    merge(g2s) %>%
    mutate(result_name = results_name)
  res_df <- bind_rows(res_df, tmp_res_df)
  
}

```

# Filtering to significant results P <0.05 and LFC abs(1)
```{r filtering DESEQ results to LFC > abs(1)& P <0.05}

filtered_res_df <- res_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

unique_sig_genes <- unique(filtered_res_df$gene_id)
```
# Testing knitting with a plot
```{R changing shapes in plot}
             


```
