---
title: "Plotting"
author: "Connor Gagen, Julia Keefe, Paloma Peters, Anna Mellizo Kroll, Anna Broerman"
date: "2025-03-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("reshape")
library(reshape)
library(tidyr)
library(tibble)
library(ggplot2)
library(dplyr)

```

```{r loading in DESeq and TPM results}
load("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/project_results/counts_DESEQ_results.rdata")

load("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/project_results/TPM_results.Rdata")
```

```{r facet all timepoint TPM distributions}
# First let's make set of time values for faceting:
time_values <- c("0_avg", "12_avg", "24_avg", "48_avg", "96_avg")

# This can also be done with grep (smarter):
time_points <- names(average_and_stddev_values)[grep("avg", names(average_and_stddev_values))]
# moving rows into df so we can track gene_id
average_and_stddev_values_2 <- rownames_to_column(average_and_stddev_values, var = "gene_id")

# Melt the data frame to long format
melt_tpm_df <- melt(average_and_stddev_values_2, measure.vars = time_points, value_name = "gene_id")

# Cool we just made it a really long df (compare original and melt) -it's 5x longer
# Note the same can be done with pivot longer
pivot_longer_tpm_df <- average_and_stddev_values_2 %>% 
  pivot_longer(cols = all_of(time_points), names_to = 'Time', values_to = 'TPM')

```

```{r TPM for DESEQ2 sig genes}
sig_genes <- as.data.frame(unique(filtered_res_df$gene_id))

names(sig_genes) <- "gene_id"

# one way to subset the data frame of avrg / sd TPM values
sig_gene_tpm <- inner_join(average_and_stddev_values_2, sig_genes, by = "gene_id")


# let's now merge in the gene names into this handy dataframe
sig_gene_tpm <- merge(sig_gene_tpm, g2s)

# Now we should organize so we can plot specific genes (not mix character and integer)
sig_gene_tpm <- sig_gene_tpm %>%
  select(-gene_id) 
```

```{r Finding top 5 significant genes}
sig_gene_tpm$change <- abs(sig_gene_tpm[[3]] - sig_gene_tpm[[1]])

# Order the data frame by the 4th and 2nd columns in ascending order (smallest values first)
sorted_df <- sig_gene_tpm[order(sig_gene_tpm[[4]], sig_gene_tpm[[2]]), ]

# Get indices of the top 5 maximum changes in the 1st and 3rd column difference
top_indices <- order(sorted_df$change, decreasing = TRUE)[1:5]

# Retrieve the top 5 values with the gene name (11th column), change, 4th column, and 2nd column
top_genes <- sorted_df[top_indices, c(11, which(colnames(sorted_df) == "change"), 4, 2)]
```
#ENSMUSG00000064357.1 is mt-Atp6
#ENSMUSG00000000031.16 is H19
#ENSMUSG00000050621.7 is Rps27rt

# (1) set of genes of interest or list.
genes_of_interest <- data.frame(gene = c("ENSMUSG00000064357.1", "ENSMUSG00000000031.16", "ENSMUSG00000050621.7"))

# (2) Reduce TPM file to these genes
df_plot_genes <- TPM_filtered[rownames(TPM_filtered) %in% genes_of_interest$gene, ]

# Moving rownames in to a column
df_plot_genes <- rownames_to_column(df_plot_genes, var = "gene_id")

# (3) Pivot longer (watch the changes !)
df_plot_genes <- df_plot_genes %>% 
  pivot_longer(cols = -gene_id, names_to = 'Time_Replicate', values_to = 'TPM')

# Seperating sample name into time and replicate for plotting table
df_plot_genes <- df_plot_genes %>%
  separate(
    Time_Replicate,
    into = c('WT', 'Time', 'Replicate'),
    sep = '_',
    extra = "merge"
  ) %>%
  select(-WT)

# Convert Time to numeric values
df_plot_genes$Time <- as.numeric(df_plot_genes$Time)

# (4) Group by Time and calculate mean and std of TPM
df_plot_genes <- df_plot_genes %>% 
  group_by(gene_id, Time) %>% 
  summarise(mean_TPM = mean(TPM), sd_TPM = sd(TPM))

# (5) Plot the line plot with error bars
ggplot(df_plot_genes, aes(x = Time, y = mean_TPM)) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_TPM - sd_TPM, ymax = mean_TPM + sd_TPM), width = 0.2) +
  labs(x = 'Time (hours)', y = 'Average TPM') +
  ggtitle('Average TPM across Time Points') +
  facet_wrap(~ gene_id, scales = "free_y")






#Plotting one specific gene: 

# Picking a favorite gene
gene_up <- as.data.frame(c("Rps27rt"))

names(gene_up) <- "gene_name"

# indexing to just get values for this gene
Rps27rt_tpm <- sig_gene_tpm[sig_gene_tpm$gene_name %in% gene_up$gene_name, ]

# now moving the character gene_name to rownames to won't interfere with numeric values
rownames(Rps27rt_tpm) <- Rps27rt_tpm$gene_name

# removing text column
Rps27rt_tpm <- Rps27rt_tpm %>%
  select(-gene_name) 
# We now need to futher prepare our data :)
# Let's get a variable for col names to plot
avg_cols <- grep("_avg", names(Rps27rt_tpm), value = TRUE)

sd_cols <- grep("_sd", names(Rps27rt_tpm), value = TRUE)

# This is another way to "melt" or pivot longer
plot_df <- data.frame(
  time = as.numeric(gsub("_avg", "", avg_cols)),
  avg = unlist(Rps27rt_tpm[avg_cols]),
  sd = unlist(Rps27rt_tpm[sd_cols])
)

# NOW THE PLOT !
ggplot(plot_df, aes(x = time, y = avg)) +
  geom_line() +
  geom_errorbar(aes(ymin = avg - sd, ymax = avg + sd), width = 0.2) +
  theme("paperwhite") +
  labs(x = "Time (hours)", y = "TPM Value") +
  ggtitle('Gene Rps27rt')
  
  
  
  
  