---
title: "Plotting"
author: "Connor Gagen, Julia Keefe, Paloma Peters, Anna Mellizo Kroll, Anna Broerman"
date: "2025-04-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
library(matrixStats)
library(knitr)
library(reshape)
```

# Goal: To generate line plots, heat map, and volcano plot of our chosen genes

### Loading in DESeq and TPM results
```{r loading in DESeq and TPM results}
load("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/project_results/counts_DESEQ_results.rdata")

load("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/juke6537/MASTER_CLASS/lessons/06_Differential_expression_analyses/project/project_results/TPM_results.Rdata")
```


## Generating line plots for each chosen gene
```{r facet all timepoint TPM distributions}
# Making a set of time values with grep:
time_points <- names(average_and_stddev_values)[grep("avg", names(average_and_stddev_values))]

# moving rows into df so we can track gene_id
average_and_stddev_values_2 <- rownames_to_column(average_and_stddev_values, var = "gene_id")

# Melt the data frame to long format
melt_tpm_df <- melt(average_and_stddev_values_2, measure.vars = time_points, value_name = "gene_id")

# Pivot longer df (not necessary but another format)
pivot_longer_tpm_df <- average_and_stddev_values_2 %>% 
  pivot_longer(cols = all_of(time_points), names_to = 'Time', values_to = 'TPM')

```

```{r TPM for DESEQ2 sig genes}
sig_genes <- as.data.frame(unique(filtered_res_df$gene_id))

names(sig_genes) <- "gene_id"

# Subsetting the data frame of avrg / sd TPM values
sig_gene_tpm <- inner_join(average_and_stddev_values_2, sig_genes, by = "gene_id")

# Merging the gene names into dataframe
sig_gene_tpm <- merge(sig_gene_tpm, g2s)

# organizing to plot specific genes (not mix character and integer)
sig_gene_tpm <- sig_gene_tpm %>%
  select(-gene_id) 
```

```{r Finding top 5 significant genes}
sig_gene_tpm$change <- abs(sig_gene_tpm[[3]] - sig_gene_tpm[[1]])

# Order the data frame by the 4th and 2nd columns in ascending order (smallest values first)
sorted_df <- sig_gene_tpm[order(sig_gene_tpm[[4]], sig_gene_tpm[[2]]), ]

# Get indices of the top 5 maximum changes in the 1st and 3rd column difference
top_indices <- order(sorted_df$change, decreasing = TRUE)[1:5]

# Retrieve the top 5 values with the gene name (11th column), change, 4th column, and 2nd column
top_genes <- sorted_df[top_indices, c(11, which(colnames(sorted_df) == "change"), 4, 2)]
```

### Plotting all three chosen genes in line plots together 
```{r all 3 genes together plotting}
# Our genes of interest
genes_of_interest <- data.frame(gene = c("ENSMUSG00000064357.1", "ENSMUSG00000000031.16", "ENSMUSG00000050621.7"))

# Reducing the TPM file to these genes 
df_plot_genes <- TPM_filtered[rownames(TPM_filtered) %in% genes_of_interest$gene, ]

# Moving rownames in to a column
df_plot_genes <- rownames_to_column(df_plot_genes, var = "gene_id")

# Pivot longer
df_plot_genes <- df_plot_genes %>% 
  pivot_longer(cols = -gene_id, names_to = 'Time_Replicate', values_to = 'TPM')

# Separating sample name into time and replicating for plotting table
df_plot_genes <- df_plot_genes %>%
  separate(
    Time_Replicate,
    into = c('WT', 'Time', 'Replicate'),
    sep = '_',
    extra = "merge"
  ) %>%
  select(-WT)

# Converting time to numeric values 
df_plot_genes$Time <- as.numeric(df_plot_genes$Time)

# Group by time, calculate mean and std of TPM
df_plot_genes <- df_plot_genes %>% 
  group_by(gene_id, Time) %>% 
  summarise(mean_TPM = mean(TPM), sd_TPM = sd(TPM))

# Creating a named vector for gene ID to gene name mapping
gene_labels <- c(
  "ENSMUSG00000064357.1" = "mt-Atp6",
  "ENSMUSG00000000031.16" = "H19",
  "ENSMUSG00000050621.7" = "Rps27rt"
)

# Modify the plot to use these names in facet_wrap
ggplot(df_plot_genes, aes(x = Time, y = mean_TPM)) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_TPM - sd_TPM, ymax = mean_TPM + sd_TPM), width = 0.2) +
  labs(x = 'Time (hours)', y = 'Average TPM') +
  ggtitle('Average TPM Across All Time Points') +
  facet_wrap(~ gene_id, scales = "free_y", labeller = labeller(gene_id = gene_labels))
```
These plots show the transcripts per million (TPM) of our genes of interest over time. These three specific genes yield interesting results as they show the most change in TPM in comparison with other genes, suggesting that doxycycline has a large effect on their expression. The Rps27rt and mt-Atp6 gene show significant increases between hours 0 and 12, while all three show further increases between hours 12 and 24. After about 48 hours, all three show decreases in TPM, but none decrease back to baseline TPM values.

### Plotting gene Rps27rt as a line plot:  
```{r Rps27rt plotting}
# Picking Rps27rt
gene_up <- as.data.frame(c("Rps27rt"))

names(gene_up) <- "gene_name"

# indexing to just get values for this gene
Rps27rt_tpm <- sig_gene_tpm[sig_gene_tpm$gene_name %in% gene_up$gene_name, ]

# Moving the character gene_name to rownames so it won't interfere with numeric values
rownames(Rps27rt_tpm) <- Rps27rt_tpm$gene_name

# removing text column
Rps27rt_tpm <- Rps27rt_tpm %>%
  select(-gene_name) 

# Getting a variable for col names to plot
avg_cols <- grep("_avg", names(Rps27rt_tpm), value = TRUE)

sd_cols <- grep("_sd", names(Rps27rt_tpm), value = TRUE)

# Pivot longer
plot_df <- data.frame(
  time = as.numeric(gsub("_avg", "", avg_cols)),
  avg = unlist(Rps27rt_tpm[avg_cols]),
  sd = unlist(Rps27rt_tpm[sd_cols])
)

# Plotting   
ggplot(plot_df, aes(x = time, y = avg)) +
  geom_line() +
  geom_errorbar(aes(ymin = avg - sd, ymax = avg + sd), width = 0.2) +
  theme("paperwhite") +
  labs(x = "Time (hours)", y = "TPM Value") +
  ggtitle('Gene Rps27rt TPM value over Time')

```  
  
### Plotting gene mt-Atp6 as a line plot:  
```{r mt-Atp6 plotting}
# Picking mt-Atp6 gene
gene_up <- as.data.frame(c("mt-Atp6"))

names(gene_up) <- "gene_name"

# indexing to just get values for this gene
mtAtp6_tpm <- sig_gene_tpm[sig_gene_tpm$gene_name %in% gene_up$gene_name, ]

# moving the character gene_name to rownames so it won't interfere with numeric values
rownames(mtAtp6_tpm) <- mtAtp6_tpm$gene_name

# removing text column
mtAtp6_tpm <- mtAtp6_tpm %>%
  select(-gene_name) 

# Getting a variable for col names to plot
avg_cols <- grep("_avg", names(mtAtp6_tpm), value = TRUE)

sd_cols <- grep("_sd", names(mtAtp6_tpm), value = TRUE)

# Another way to "melt" or pivot longer
plot_df <- data.frame(
  time = as.numeric(gsub("_avg", "", avg_cols)),
  avg = unlist(mtAtp6_tpm[avg_cols]),
  sd = unlist(mtAtp6_tpm[sd_cols])
)

# Plotting
ggplot(plot_df, aes(x = time, y = avg)) +
  geom_line() +
  geom_errorbar(aes(ymin = avg - sd, ymax = avg + sd), width = 0.2) +
  theme("paperwhite") +
  labs(x = "Time (hours)", y = "TPM Value") +
  ggtitle('Gene mt-Atp6 Across Time')
```  

### Plotting gene H19 as a line plot:  
```{r H19 plotting}
# Picking a favorite gene
gene_up <- as.data.frame(c("H19"))

names(gene_up) <- "gene_name"

# indexing to just get values for this gene
H19_tpm <- sig_gene_tpm[sig_gene_tpm$gene_name %in% gene_up$gene_name, ]

# now moving the character gene_name to rownames so it won't interfere with numeric values
rownames(H19_tpm) <- H19_tpm$gene_name

# removing text column
H19_tpm <- H19_tpm %>%
  select(-gene_name) 

# Getting a variable for col names to plot
avg_cols <- grep("_avg", names(H19_tpm), value = TRUE)

sd_cols <- grep("_sd", names(H19_tpm), value = TRUE)

# Another way to "melt" or pivot longer
plot_df <- data.frame(
  time = as.numeric(gsub("_avg", "", avg_cols)),
  avg = unlist(H19_tpm[avg_cols]),
  sd = unlist(H19_tpm[sd_cols])
)

# Plotting
ggplot(plot_df, aes(x = time, y = avg)) +
  geom_line() +
  geom_errorbar(aes(ymin = avg - sd, ymax = avg + sd), width = 0.2) +
  theme("paperwhite") +
  labs(x = "Time (hours)", y = "TPM Value") +
  ggtitle('Gene H19 Across Time')
  
```

## Creating a volcano plot of the chosen genes 
```{r volcano plot}
highlighted_genes <- c("H19", "Rps27rt", "mtAtp6")

# Create a new column for coloring, using the gene column 
filtered_res_df$color_group <- ifelse(filtered_res_df[[8]] %in% highlighted_genes, "Genes of Interest", "Other")

# Plot
ggplot(filtered_res_df, aes(x = log2FoldChange, y = -log10(padj), color = color_group)) +
  geom_point(alpha = 0.5) +  # Adjust transparency for clarity
  xlim(-5, 5) +
  ylim(0, 65) +
  scale_color_manual(values = c("Genes of Interest" = "red", "Other" = "black")) +  # Set colors
  labs(x = "Log2 Fold Change", y = "-log10(Adjusted P-value)", title = "Volcano Plot of Significant Genes") +
  theme_minimal()

```
  
## Creating a heatmap of the chosen genes
```{r distance calculation}

# Log-transform the TPM values. 
log_tpm_matrix <- log2(TPM_filtered + 1)

# euclidean distance / similarity based on log2 TPM values of genes in TPM_filtered.
distance_matrix <- dist(t(log_tpm_matrix), method = "euclidean")

hc_complete <- hclust(distance_matrix, method = "complete")

# use plot() to plot the distance calucalated in hclust
plot(hc_complete, labels = colnames(log_tpm_matrix), main = "Global TPM Filtered Dendrogram ", sub = "sample relationship by time and replicate after dox treatment")

# Cut tree to limit the clusters 
clusters <- cutree(hc_complete, k = 5)
```

```{r ordered dendrogram}
# Making a dendrogram of hc_complete
dend_hc_complete <- as.dendrogram(hc_complete)

# Reorder the dendrogram based on the clusters created above
dend_hc_complete <- reorder(dend_hc_complete, clusters)

# Plot the reordered dendrogram
plot(dend_hc_complete, main = "Global TPM Filtered Dendrogram (Ordered by Cluster)", sub = "Sample relationship by time and replicate after dox treatment")
```

```{r Create Heat map of all filtered TPM genes across all time points and replicates}
# Mapping gene IDs to gene names
gene_labels <- c(
  "ENSMUSG00000064357.1" = "mt-Atp6",
  "ENSMUSG00000000031.16" = "H19",
  "ENSMUSG00000050621.7" = "Rps27rt"
)

# Subset the log_tpm_matrix to include only the genes of interest using their gene IDs
log_tpm_subset <- log_tpm_matrix[rownames(log_tpm_matrix) %in% names(gene_labels), ]

# Rename row names to gene names using gene_labels
rownames(log_tpm_subset) <- gene_labels[rownames(log_tpm_subset)]


custom_dist <- function(x) dist(x, method = "manhattan") # Replace "manhattan" with your preferred method
custom_hclust <- function(x) hclus


# Create the heatmap with the subset of genes for the time points
deseq_samples$time_point <- factor(deseq_samples$time_point, levels = c("0", "12", "24", "48", "96"))

# reorder the deseq samples in sequential order
deseq_samples <- deseq_samples[order(deseq_samples$time_point), ]

# Reorders the tpm based on the new order
log_tpm_subset_ordered <- log_tpm_subset[, deseq_samples$sample_id]

# Generate the heatmap
pheatmap(log_tpm_subset_ordered, 
         cluster_rows = TRUE,  
         cluster_cols = FALSE,  # Disable clustering for columns to preserve order
         scale = "row",        
         show_rownames = TRUE, 
         show_colnames = TRUE,  
         main = "Heatmap of Log2-Transformed TPM Values for Selected Genes",
         color = colorRampPalette(c("blue", "white", "red"))(50))

```
